---
title: "Lab 4 Assignment - Landcover"
author: "Justin Hall"
date: "2-12-2024"
output:
  html_document:
    df_print: paged
---


```{r, warning=F, error=F, message=F}

# Clear Workspace
rm(list=ls())

# Load Packages
require(Voss)
require(tidyverse)
require(terra)
require(FedData)
require(sf)
require(tidyterra)
require(landscapemetrics)
require(ggplot2)
```


## Challenge 1 (4 points)

**The landscapemetrics package has functions for calculating 12 patch-level metrics. Calculate all 12 of these for every forest patch in our nlcdSimple raster (using the 8-direction rule). This will result in 12 unique values associated with each forest patch. Use the chart.Correlation() function from the PerformanceAnalytics package to examine the correlation among all 12 of these variables. What patterns do you notice, and what do they tell you about the uniqueness or redundancy of the patch-level metrics you calculated?**

```{r, warning=F, error=F, message=F}

# Bring In Study Area
studyArea = st_read('/vsicurl/https://github.com/ValenteJJ/SpatialEcology/raw/main/Week4/studyArea.shp')

# Bring in NLCD Data for Study Area
nlcd = get_nlcd(studyArea, label='AlLandscape', year=2011)

# Simplify NLCD Categories
nlcdSimple = nlcd
nlcdSimple[nlcdSimple==11] = 1 #Wet areas are a 1 now
nlcdSimple[nlcdSimple %in% c(21, 22, 23, 24)] = 2 #All developed areas are 2
nlcdSimple[nlcdSimple %in% c(31, 52)] = 3 #Barren land and shrub/scrub are 3
nlcdSimple[nlcdSimple %in% c(41,42,43)] = 4 #All forest types are 4
nlcdSimple[nlcdSimple == 71] = 5 #Grassland is 5
nlcdSimple[nlcdSimple %in% c(81,82)] = 6 #And agriculture is 6

# Define Reclassified Values with Labels
tmp = data.frame('ID' = c(1, 2, 3, 4, 5, 6),
                 'category' = c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture'))
nlcdSimple = categories(nlcdSimple, value=tmp)

# Plot the Simplified Raster
ggplot(nlcdSimple, aes(x=x, y=y, fill=category)) +
  geom_raster()+
  scale_fill_manual(values=c('blue', 'black', 'gray', 'darkolivegreen', 'orange', 'yellow'))
  
# Convert Simplifed Raster to Forest Cover
forest = nlcdSimple %>% 
setValues(NA)
forest[nlcdSimple ==4] = 1
forestPatchId = patches(forest, directions=8, zeroAsNA=T, allowGaps=F)

# Plot Forest Cover
plot(forestPatchId)
```

```{r, warning=F, error=F, message=F}

# Calculate All Patch-level Metrics

# Area
patchArea = lsm_p_area(forest, directions=8)
patchArea

# Core area index
patchcai = lsm_p_cai(forest, directions=8)
patchcai

# Circle
patchcircle = lsm_p_circle(forest, directions=8)
patchcircle

# Contiguity index
patchcontig = lsm_p_contig(forest, directions=8)
patchcontig

# Core
patchCore = lsm_p_core(forest, directions=8)
patchCore

# Euclidean Nearest-Neighbor Distance
patchEnn = lsm_p_enn(forest, directions=8)
patchEnn

# Fractal dimension index
patchFrac = lsm_p_frac(forest, directions=8)
patchFrac

# Radius of Gyration
patchGyrate = lsm_p_gyrate(forest, directions=8)
patchGyrate

# Number of core areas
patchNcore = lsm_p_ncore(forest, directions=8)
patchNcore

# Perimeter-Area ratio
patchPara = lsm_p_para(forest, directions=8)
patchPara

# Perimeter
patchPer = lsm_p_perim(forest, directions=8)
patchPer

# Shape
patchShape = lsm_p_shape(forest, directions=8)
patchShape

  
```

```{r, warning=F, error=F, message=F}

# Combine Patch-level Metric Values Into a Single Data Frame
landscape_metrics_df <- data.frame(
  Area = patchArea$value,
  CoreAreaIndex = patchcai$value,
  Circle = patchcircle$value,
  ContiguityIndex = patchcontig$value,
  Core = patchCore$value,
  EuclideanNNDistance = patchEnn$value,
  FractalDimensionIndex = patchFrac$value,
  RadiusOfGyration = patchGyrate$value,
  NumberOfCoreAreas = patchNcore$value,
  PerimeterAreaRatio = patchPara$value,
  Perimeter = patchPer$value,
  Shape = patchShape$value
)

# Examine the Correlations Among Patch-level Metrics
PerformanceAnalytics::chart.Correlation(landscape_metrics_df, histogram=TRUE, pch=19)
  
```

**ANSWER:** 
The correlation plots reveal diverse patterns among the 12 patch-level metrics, emphasizing highly significant correlations. Some metrics like "Area" and "Core" are redundant, with a correlation coefficient of 1.00, while "Contiguity" and "Perimeter-area ratio" show a near-perfect inverse correlation of -0.99. Other metrics also demonstrate significant relationships, with some showing no significant correlation, indicating potential redundancy in further analysis.


## Challenge 2 (4 points)

**In our lab, we used the 8-direction or "queen" rule to delineate patches. Using the nlcdSimple raster we created, explore the differences in patch characteristics if you were to use the 4-direction or "rook" rule for delineating patches. Calculate the following class-level metrics for forest cover for both the queen and rook patch delineation rules: number of patches, mean patch size, standard deviation in patch size, nearest-neighbor distance, and total edge length. What patterns do you notice? When might it be appropriate to use one rule vs. the other?**

```{r, warning=F, error=F, message=F}

# Queen Delineation Rules (8 Directions)

# Number of Patches
NumPatch8 = lsm_c_np(forest, directions=8)
NumPatch8

# Mean Patch Size
MeanPatch8 = lsm_c_area_mn(forest, directions=8)
MeanPatch8

# Standard Deviation in Patch Size
SDPatch8 = lsm_c_area_sd(forest, directions=8)
SDPatch8

# Mean Nearest-neigbor Distance
NNDistance8 = lsm_c_enn_mn(forest, directions=8)
NNDistance8

################################################################################
# --> Create New Forest Layer for Computing Total Edge <--
forest0 = forest
forest0[is.na(forest)]=0
################################################################################

# Total Edge Length
TEdge8 = lsm_c_te(forest0, directions=8)
TEdge8



#Rook Delineation Rules (4 Directions)

# Number of Patches
NumPatch4 = lsm_c_np(forest, directions=4)
NumPatch4

# Mean Patch Size
MeanPatch4 = lsm_c_area_mn(forest, directions=4)
MeanPatch4

# Standard Deviation in Patch Size
SDPatch4 = lsm_c_area_sd(forest, directions=4)
SDPatch4

# Mean Nearest-neigbor Distance
NNDistance4 = lsm_c_enn_mn(forest, directions=4)
NNDistance4

# Total Edge Length --> Again, use forest0 to avoid total edge value = 0
TEdge4 = lsm_c_te(forest0, directions=4)
TEdge4
```

**ANSWER:** 
Queen connectivity (8 directions) leads to fewer and larger patches, while Rook connectivity (4 directions) generates more but smaller patches. Higher standard deviation in patch size under Queen connectivity likely results from the contrast between larger patches and smaller specks on the landscape. Closer nearest neighbor patches under Rook connectivity may be due to the increased number of patches. Total edge length remains consistent under both connectivity levels, indicating that the perimeter of non-habitat areas remains unchanged regardless of connectivity rules.

The species' biology is critical when deciding between Queen and Rook connectivity rules. High-mobility species benefit from Queen rules, facilitating movement between diagonally connected habitats. Conversely, low-mobility species may prefer Rook rules, as small gaps between larger habitat chunks may be inaccessible.


## Challenge 3 (4 points)


**Using the same zoomed-in study area that we used in the lab, download NLCD raster data for the years 2001 and 2019 and simplify these rasters into 6 landcover categories (wet, developed, barren, forest, grassland, and agriculture). Plot these two rasters. What are some of the initial changes you notice between 2001 and 2019?**

```{r, warning=F, error=F, message=F}
# Pull NLCD Data For Both Years
nlcd2001 = get_nlcd(studyArea, label='AlLandscape', year=2001)
nlcd2019 = get_nlcd(studyArea, label='AlLandscape', year=2019)

# Simplify 2001
nlcd2001 = nlcd2001
nlcd2001[nlcd2001==11] = 1 #Wet areas are a 1 now
nlcd2001[nlcd2001 %in% c(21, 22, 23, 24)] = 2 #All developed areas are 2
nlcd2001[nlcd2001 %in% c(31, 52)] = 3 #Barren land and shrub/scrub are 3
nlcd2001[nlcd2001 %in% c(41,42,43)] = 4 #All forest types are 4
nlcd2001[nlcd2001 == 71] = 5 #Grassland is 5
nlcd2001[nlcd2001 %in% c(81,82)] = 6 #And agriculture is 6

# Define Reclassified Values with Labels
tmp = data.frame('ID' = c(1, 2, 3, 4, 5, 6),
                 'category' = c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture'))
nlcd2001 = categories(nlcd2001, value=tmp)


# Simplify 2019
nlcd2019 = nlcd2019
nlcd2019[nlcd2019==11] = 1 # Wet areas are a 1 now
nlcd2019[nlcd2019 %in% c(21, 22, 23, 24)] = 2 # All developed areas are 2
nlcd2019[nlcd2019 %in% c(31, 52)] = 3 # Barren land and shrub/scrub are 3
nlcd2019[nlcd2019 %in% c(41,42,43)] = 4 # All forest types are 4
nlcd2019[nlcd2019 == 71] = 5 # Grassland is 5
nlcd2019[nlcd2019 %in% c(81,82)] = 6 # And agriculture is 6

# Define Reclassified Values with Labels
tmp = data.frame('ID' = c(1, 2, 3, 4, 5, 6),
                 'category' = c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture'))
nlcd2019 = categories(nlcd2019, value=tmp) # This line may need adjustment based on actual function to assign categories



# Plot 2001 and 2019 NLCD data
# 2001
ggplot(nlcd2001, aes(x=x, y=y, fill=category)) +
  geom_raster()+
  scale_fill_manual(values=c('blue', 'black', 'gray', 'darkolivegreen', 'orange', 'yellow'))

# 2019
ggplot(nlcd2019, aes(x=x, y=y, fill=category)) +
  geom_raster() +
  scale_fill_manual(values=c('blue', 'black', 'gray', 'darkolivegreen', 'orange', 'yellow'))



```

**ANSWER:** 
Between 2001 and 2019, significant forest loss occurred, replaced by grassland, agriculture, and open areas, likely due to agricultural expansion.

**Quantify this at the class level by calculating and reporting the changes in (1) the total amount of each land cover type (2) mean patch size for each land cover type, and (3) mean nearest neighbor distance for each cover type between the years 2011 and 2019. Give a short description of how you interpret the changes in these values.**

```{r, warning=F, error=F, message=F}
# Total Amount of Each Land Cover Type

# 2001
# Calculation
TotArea2001 = lsm_c_ca(nlcd2001, directions=8)

# Change numeric classes to category names
class_labels <- c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture')
TotArea2001$class <- factor(TotArea2001$class, levels = 1:6, labels = class_labels)
TotArea2001

# 2019
# Calculation
TotArea2019 = lsm_c_ca(nlcd2019, directions=8)

# Change numeric classes to category names
class_labels <- c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture')
TotArea2019$class <- factor(TotArea2019$class, levels = 1:6, labels = class_labels)
TotArea2019

# Difference
Area_Change <- TotArea2019$value - TotArea2001$value
names(Area_Change) <- class_labels
Area_Change
################################################################################

# Mean Patch Size of Each Land Cover Type

# 2001
# Calculation
MeanPatch2001 = lsm_c_area_mn(nlcd2001, directions=8)

# Change numeric classes to category names
class_labels <- c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture')
MeanPatch2001$class <- factor(MeanPatch2001$class, levels = 1:6, labels = class_labels)
MeanPatch2001

# 2019
# Calculation
MeanPatch2019 = lsm_c_area_mn(nlcd2019, directions=8)

# Change numeric classes to category names
class_labels <- c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture')
MeanPatch2019$class <- factor(MeanPatch2019$class, levels = 1:6, labels = class_labels)
MeanPatch2019

# Difference
MeanPatch_Change <- MeanPatch2019$value - MeanPatch2001$value
names(MeanPatch_Change) <- class_labels
MeanPatch_Change
################################################################################

# Mean Nearest Neighbor Distance of Each Land Cover Type

# 2001
# Calculation
MeanNND2001 = lsm_c_enn_mn(nlcd2001, directions=8)

# Change numeric classes to category names
class_labels <- c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture')
MeanNND2001$class <- factor(MeanNND2001$class, levels = 1:6, labels = class_labels)
MeanNND2001

# 2019
# Calculation
MeanNND2019 = lsm_c_enn_mn(nlcd2019, directions=8)

# Change numeric classes to category names
class_labels <- c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture')
MeanNND2019$class <- factor(MeanNND2019$class, levels = 1:6, labels = class_labels)
MeanNND2019

# Difference
MeanNND_Change <- MeanNND2019$value - MeanNND2001$value
names(MeanNND_Change) <- class_labels
MeanNND_Change
```

**ANSWER:** As demonstrated by the labeled "Change" vectors, the following metrics have been changed in the following ways between 2001 and 2019:

*(1) Total amount of each land cover type:*

- Wetland: No Change
- Developed: Gain of 1.35 hectares
- Open: Gain of 159.39 hectares
- Forest: Loss of 413.46 hectares
- Grassland: Gain of 125.64 hectares
- Agriculture: Gain of 127.08 hectares

*The takeaway in this scenario is that forest has been lost and replaced by agriculture, open areas, and grassland habitat types.*


*(2) Mean patch size for each land cover type*

- Wetland: No Change
- Developed: Loss of 0.18 hectares
- Open: Gain of 2.53 hectares
- Forest: Loss of 31.16 hectares
- Grassland: Gain of 1.77 hectares
- Agriculture: Gain of 4.78 hectares

*The takeaway in this scenario is that average patch size for most of the land cover types has increased with the decreasing forest patch sizes (on average).* 

**(3) Mean nearest neighbor distance for each cover type**

- Wetland: No Change
- Developed: Loss of 6.42 meters
- Open: Loss of 197.80 meters
- Forest: Loss of 4.26 meters
- Grassland: Loss of 64.44 meters
- Agriculture: Loss of 3.12 meters

*The takeaway in this scenario is that mean nearest neighbor patches decreased for all land cover types except wetlands. This indicates that patches are, on average, closer to each other, likely due to fragmentation.*

**Quantify these changes at the landscape level by calculating and reporting on changes in the (1) Shannon diversity and (2) Shannon evenness of the landscapes at the different time points. Give a short description of how you interpret the changes in these values.** 

```{r, warning=F, error=F, message=F}

# Shannon Diversity

# 2001
# Calculation
SDiversity2001 = lsm_l_shdi(nlcd2001)
SDiversity2001

# 2019
# Calculation
SDiversity2019 = lsm_l_shdi(nlcd2019)
SDiversity2019

# Difference
SDiversity_Change <- SDiversity2019$value - SDiversity2001$value
SDiversity_Change


# Shannon Evenness

# 2001
# Calculation
SEvenness2001 = lsm_l_shei(nlcd2001)
SEvenness2001

# 2019
# Calculation
SEvenness2019 = lsm_l_shei(nlcd2019)
SEvenness2019

# Difference
SEvenness_Change <- SEvenness2019$value - SEvenness2001$value
SEvenness_Change


```

**ANSWER:** The provided code shows that from 2011 to 2019, the Shannon's Diversity Index for landscape classes increased by 0.3930831, and Shannon's evenness index increased by 0.2193839, indicating heightened landscape diversity with a more balanced representation of land cover types. Despite no entire class gains or losses among the observed six land cover types, the increased diversity isn't necessarily beneficial for wildlife, as certain cover types like agriculture and developed areas often provide poor habitat. Additionally, the fragmentation of land cover types can have diverse effects on wildlife communities, depending on species-specific adaptations.


## Challenge 4 (4 points)

**Use the voss2d() function to simulate a surface where g = 7 and H = 0.5. From that, create 9 'landscapes' with 10%, 20%, 30%, ..., 90% threshold values. The '1' values here can represent anything your imagination comes up with. It could be forest cover, cover of some other land cover type, bodies of water, temperatures above a threshold, etc. I suggest you set the seed for your simulation value so that you get the same outcome each time you run the code. Plot these landscapes and comment on what patterns you see changing as the value increases from a 10% cover to 90% cover.**

```{r, warning=F, error=F, message=F}
set.seed(23)

#Set vossModel Parameters
vossModel = voss2d(g=7, H=0.5)
vossModel = rast(vossModel$z)

# Plot Initial vossModel
plot(vossModel)

# 10% Forest cover
threshold10 = quantile(as.matrix(vossModel), prob=0.1)
voss10 = ifel(vossModel > threshold10, 0, 1)
plot(voss10)

# 20% Forest cover
threshold20 = quantile(as.matrix(vossModel), prob=0.2)
voss20 = ifel(vossModel > threshold20, 0, 1)
plot(voss20)

# 30% Forest cover
threshold30 = quantile(as.matrix(vossModel), prob=0.3)
voss30 = ifel(vossModel > threshold30, 0, 1)
plot(voss30)

# 40% Forest cover
threshold40 = quantile(as.matrix(vossModel), prob=0.4)
voss40 = ifel(vossModel > threshold40, 0, 1)
plot(voss40)

# 50% Forest cover
threshold50 = quantile(as.matrix(vossModel), prob=0.5)
voss50 = ifel(vossModel > threshold50, 0, 1)
plot(voss50)

# 60% Forest cover
threshold60 = quantile(as.matrix(vossModel), prob=0.6)
voss60 = ifel(vossModel > threshold60, 0, 1)
plot(voss60)

# 70% Forest cover
threshold70 = quantile(as.matrix(vossModel), prob=0.7)
voss70 = ifel(vossModel > threshold70, 0, 1)
plot(voss70)

# 80% Forest cover
threshold80 = quantile(as.matrix(vossModel), prob=0.8)
voss80 = ifel(vossModel > threshold80, 0, 1)
plot(voss80)

# 90% Forest cover
threshold90 = quantile(as.matrix(vossModel), prob=0.9)
voss90 = ifel(vossModel > threshold90, 0, 1)
plot(voss90)

```

**ANSWER:** As forest cover expands, it extends mainly from existing forested areas, reducing the total number of distinct forest patches as they connect. This typically increases the average patch size and reduces distances to nearest neighboring patches due to more overall forest cover nearby.

**Identify 3 class-level or landscape-level metrics that help you capture the changes you are observing. Calculate those metrics for each of the 9 landscapes and plot them (x-axis is threshold value and y-axis is calculated metric). Briefly describe why you chose these 3 metrics and how they change with increasing cover.**

```{r, warning=F, error=F, message=F}

# Number of Patches
NumPatchVoss10 = lsm_c_np(voss10, directions=8)
NumPatchVoss10 <- NumPatchVoss10$value[2]
NumPatchVoss20 = lsm_c_np(voss20, directions=8)
NumPatchVoss20 <- NumPatchVoss20$value[2]
NumPatchVoss30 = lsm_c_np(voss30, directions=8)
NumPatchVoss30 <- NumPatchVoss30$value[2]
NumPatchVoss40 = lsm_c_np(voss40, directions=8)
NumPatchVoss40 <- NumPatchVoss40$value[2]
NumPatchVoss50 = lsm_c_np(voss50, directions=8)
NumPatchVoss50 <- NumPatchVoss50$value[2]
NumPatchVoss60 = lsm_c_np(voss60, directions=8)
NumPatchVoss60 <- NumPatchVoss60$value[2]
NumPatchVoss70 = lsm_c_np(voss70, directions=8)
NumPatchVoss70 <- NumPatchVoss70$value[2]
NumPatchVoss80 = lsm_c_np(voss80, directions=8)
NumPatchVoss80 <- NumPatchVoss80$value[2]
NumPatchVoss90 = lsm_c_np(voss90, directions=8)
NumPatchVoss90 <- NumPatchVoss90$value[2]


# Data Frame to Plot
NumPatchData <- data.frame(
  Threshold = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9),
  NumPatches = c(NumPatchVoss10, NumPatchVoss20, NumPatchVoss30,
                 NumPatchVoss40, NumPatchVoss50, NumPatchVoss60,
                 NumPatchVoss70, NumPatchVoss80, NumPatchVoss90)
)

# Create Line of Best Fit For Inclusion on Figure
PatchModel <- lm(NumPatches ~ Threshold, data = NumPatchData)
PatchSlope <- coef(PatchModel)["Threshold"]
PatchIntercept <- coef(PatchModel)["(Intercept)"]
PatchLineLabel <- paste("y =", round(PatchIntercept, 2), round(PatchSlope, 2), "x")



ggplot(NumPatchData, aes(x = Threshold, y = NumPatches)) +
  geom_point() +  
  geom_smooth(method = "lm", se = TRUE, color = "red") + 
  annotate("text", x = 0.5, y = min(NumPatchData$NumPatches) * 1.05, label = PatchLineLabel, hjust = 0, vjust = 0, color = "red") +
  theme_minimal() +
  labs(title = "Number of Patches vs Threshold Values",
       x = "Threshold",
       y = "Number of Patches")

################################################################################

# Mean Patch Size
MeanPatch10 = lsm_c_area_mn(voss10, directions=8)
MeanPatch10 <- MeanPatch10$value[2]
MeanPatch20 = lsm_c_area_mn(voss20, directions=8)
MeanPatch20 <- MeanPatch20$value[2]
MeanPatch30 = lsm_c_area_mn(voss30, directions=8)
MeanPatch30 <- MeanPatch30$value[2]
MeanPatch40 = lsm_c_area_mn(voss40, directions=8)
MeanPatch40 <- MeanPatch40$value[2]
MeanPatch50 = lsm_c_area_mn(voss50, directions=8)
MeanPatch50 <- MeanPatch50$value[2]
MeanPatch60 = lsm_c_area_mn(voss60, directions=8)
MeanPatch60 <- MeanPatch60$value[2]
MeanPatch70 = lsm_c_area_mn(voss70, directions=8)
MeanPatch70 <- MeanPatch70$value[2]
MeanPatch80 = lsm_c_area_mn(voss80, directions=8)
MeanPatch80 <- MeanPatch80$value[2]
MeanPatch90 = lsm_c_area_mn(voss90, directions=8)
MeanPatch90 <- MeanPatch90$value[2]

# Data Frame to Plot
MeanPatchData <- data.frame(
  Threshold = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9),
  MeanPatches = c(MeanPatch10, MeanPatch20, MeanPatch30,
                 MeanPatch40, MeanPatch50, MeanPatch60,
                 MeanPatch70, MeanPatch80, MeanPatch90)
)

# Create Line of Best Fit For Inclusion on Figure
MeanPatchModel <- lm(MeanPatches ~ Threshold, data = MeanPatchData)
MeanPatchSlope <- coef(MeanPatchModel)["Threshold"]
MeanPatchIntercept <- coef(MeanPatchModel)["(Intercept)"]
MeanPatchLineLabel <- paste("y =", round(MeanPatchIntercept, 2), "+", round(MeanPatchSlope, 2), "x")


ggplot(MeanPatchData, aes(x = Threshold, y = MeanPatches)) +
  geom_point() +  
  geom_smooth(method = "lm", se = TRUE, color = "red") + 
  annotate("text", x = 0.5, y = min(MeanPatchData$MeanPatches) * 1.05, label = MeanPatchLineLabel, hjust = 0, vjust = 0, color = "red") +
  theme_minimal() +
  labs(title = "Mean Patch Size vs Threshold Value",
       x = "Threshold",
       y = "Mean Patch Size")
################################################################################

# Mean Nearest-neigbor Distance

ENN10 = lsm_c_enn_mn(voss10, directions=8)
ENN10 <- ENN10$value[2]
ENN20 = lsm_c_enn_mn(voss20, directions=8)
ENN20 <- ENN20$value[2]
ENN30 = lsm_c_enn_mn(voss30, directions=8)
ENN30 <- ENN30$value[2]
ENN40 = lsm_c_enn_mn(voss40, directions=8)
ENN40 <- ENN40$value[2]
ENN50 = lsm_c_enn_mn(voss50, directions=8)
ENN50 <- ENN50$value[2]
ENN60 = lsm_c_enn_mn(voss60, directions=8)
ENN60 <- ENN60$value[2]
ENN70 = lsm_c_enn_mn(voss70, directions=8)
ENN70 <- ENN70$value[2]
ENN80 = lsm_c_enn_mn(voss80, directions=8)
ENN80 <- ENN80$value[2]
ENN90 = lsm_c_enn_mn(voss90, directions=8)
ENN90 <- ENN90$value[2]


# Data Frame to Plot
ENNData <- data.frame(
  Threshold = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9),
  ENNValues = c(ENN10, ENN20, ENN30,
                 ENN40, ENN50, ENN60,
                 ENN70, ENN80, ENN90)
)

# Create Line of Best Fit For Inclusion on Figure
ENNModel <- lm(ENNValues ~ Threshold, data = ENNData)
ENNSlope <- coef(ENNModel)["Threshold"]
ENNIntercept <- coef(ENNModel)["(Intercept)"]
ENNLineLabel <- paste("y =", round(ENNIntercept, 2), round(ENNSlope, 2), "x")

ggplot(ENNData, aes(x = Threshold, y = ENNValues)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red") + 
  annotate("text", x = 0.5, y = min(ENNData$ENNValues) * 1.05, label = ENNLineLabel, hjust = 0, vjust = 0, color = "red") +
  theme_minimal() +
  labs(title = "Mean Nearest-Neighbor Distance vs Threshold Value",
       x = "Threshold",
       y = "Mean Nearest-Neighbor Distance")

```

**Identify 3 class-level or landscape-level metrics that help you capture the changes you are observing. Calculate those metrics for each of the 9 landscapes and plot them (x-axis is threshold value and y-axis is calculated metric). Briefly describe why you chose these 3 metrics and how they change with increasing cover.**

**ANSWER:** 
*Class Level Nearest-Neighbor Mean Distance:* This metric was chosen because it is anticipated that increasing cover would expand spatial coverage, reduce open space, and decrease the average distance between patches. This was supported by the negative slope observed in the comparison of threshold to mean nearest-neighbor values. Among the three metrics studied, this one appears the most linear to me. Typically, with low tree cover, it is expected to be sparse forest patches separated by long distances. Conversely, as forest patches saturate, it is anticipated a steady decline in the distance between them.

*Class Level Mean Patch Size:* This metric was chosen because forest cover appeared to increase systematically, leading to the prediction that mean patch size would also increase with threshold as patches connected. While plotting threshold against mean patch size revealed a positive slope, the true relationship is likely nonlinear.

*Class Level Number of Patches:* This metric was chosen because as forest cover increased using the Voss package, it systematically expanded from existing forest patches. It is anticipated that as forest cover expanded, patches would connect, reducing the total number of distinct patches. When plotting threshold against the number of patches, a visual assessment suggested a negative slope. However, since 0% cover results in 0 patches and 100% cover results in 1 patch, the relationship may not be best represented by a linear slope.


## Challenge 5 (4 points)

**Use the voss2d() function to simulate 9 surfaces. In each one, g should be 7, but vary the value of H from 0.1 to 0.9. Create a unique landscape from each of these with a threshold value of 30% cover. Again, the 1 values can represent anything you desire, and again I suggest you set the seed for your simulation value so that you get the same outcome each time you run the code. Plot these landscapes and comment on what patterns you see changing as the H value increases from 0.1 to 0.9.**

```{r, warning=F, error=F, message=F}
set.seed(23)

# Set Parameters for vossModels
vossModel.1 = voss2d(g=7, H=0.1)
vossModel.1 = rast(vossModel.1$z)

vossModel.2 = voss2d(g=7, H=0.2)
vossModel.2 = rast(vossModel.2$z)

vossModel.3 = voss2d(g=7, H=0.3)
vossModel.3 = rast(vossModel.3$z)

vossModel.4 = voss2d(g=7, H=0.4)
vossModel.4 = rast(vossModel.4$z)

vossModel.5 = voss2d(g=7, H=0.5)
vossModel.5 = rast(vossModel.5$z)

vossModel.6 = voss2d(g=7, H=0.6)
vossModel.6 = rast(vossModel.6$z)

vossModel.7 = voss2d(g=7, H=0.7)
vossModel.7 = rast(vossModel.7$z)

vossModel.8 = voss2d(g=7, H=0.8)
vossModel.8 = rast(vossModel.8$z)

vossModel.9 = voss2d(g=7, H=0.9)
vossModel.9 = rast(vossModel.9$z)


# Plot voss Forest Covers

# 30% Forest Cover --> H = 0.1
surface1 = quantile(as.matrix(vossModel.1), prob=0.3)
vosssurface1 = ifel(vossModel.1 > surface1, 0, 1)
plot(vosssurface1)

# 30% Forest Cover --> H = 0.2
surface2 = quantile(as.matrix(vossModel.2), prob=0.3)
vosssurface2 = ifel(vossModel.2 > surface2, 0, 1)
plot(vosssurface2)

# 30% Forest Cover --> H = 0.3
surface3 = quantile(as.matrix(vossModel.3), prob=0.3)
vosssurface3 = ifel(vossModel.3 > surface3, 0, 1)
plot(vosssurface3)

# 30% Forest Cover --> H = 0.4
surface4 = quantile(as.matrix(vossModel.4), prob=0.3)
vosssurface4 = ifel(vossModel.4 > surface4, 0, 1)
plot(vosssurface4)

# 30% Forest Cover --> H = 0.5
surface5 = quantile(as.matrix(vossModel.5), prob=0.3)
vosssurface5 = ifel(vossModel.5 > surface5, 0, 1)
plot(vosssurface5)

# 30% Forest Cover --> H = 0.6
surface6 = quantile(as.matrix(vossModel.6), prob=0.3)
vosssurface6 = ifel(vossModel.6 > surface6, 0, 1)
plot(vosssurface6)

# 30% Forest Cover --> H = 0.7
surface7 = quantile(as.matrix(vossModel.7), prob=0.3)
vosssurface7 = ifel(vossModel.7 > surface7, 0, 1)
plot(vosssurface7)

# 30% Forest Cover --> H = 0.8
surface8 = quantile(as.matrix(vossModel.8), prob=0.3)
vosssurface8 = ifel(vossModel.8 > surface8, 0, 1)
plot(vosssurface8)

# 30% Forest Cover --> H = 0.9
surface9 = quantile(as.matrix(vossModel.9), prob=0.3)
vosssurface9 = ifel(vossModel.9 > surface9, 0, 1)
plot(vosssurface9)

```

**ANSWER:** As H increases from 0.1 to 0.9, forest cover becomes more clumped and less patchy. At H = 0.9, it appears as one large patch with isolated pixels, while at H = 0.1, it's significantly more scattered across the landscape.


**Identify 3 class-level or landscape-level metrics that help you capture the changes you are observing. THESE MUST BE DIFFERENT THAN THOSE METRICS YOU USED IN CHALLENGE 2. Calculate those metrics for each of the 9 landscapes and plot them (x-axis is H-value and y-axis is calculated metric). Briefly describe why you chose these 3 metrics and how they change with increasing cover.**

```{r, warning=F, error=F, message=F}
# Landscape Patch Cohesion
PatchCohesion.1 = lsm_l_cohesion(vosssurface1, directions=8)
PatchCohesion.1 <- PatchCohesion.1$value
PatchCohesion.2 = lsm_l_cohesion(vosssurface2, directions=8)
PatchCohesion.2 <- PatchCohesion.2$value
PatchCohesion.3 = lsm_l_cohesion(vosssurface3, directions=8)
PatchCohesion.3 <- PatchCohesion.3$value
PatchCohesion.4 = lsm_l_cohesion(vosssurface4, directions=8)
PatchCohesion.4 <- PatchCohesion.4$value
PatchCohesion.5 = lsm_l_cohesion(vosssurface5, directions=8)
PatchCohesion.5 <- PatchCohesion.5$value
PatchCohesion.6 = lsm_l_cohesion(vosssurface6, directions=8)
PatchCohesion.6 <- PatchCohesion.6$value
PatchCohesion.7 = lsm_l_cohesion(vosssurface7, directions=8)
PatchCohesion.7 <- PatchCohesion.7$value
PatchCohesion.8 = lsm_l_cohesion(vosssurface8, directions=8)
PatchCohesion.8 <- PatchCohesion.8$value
PatchCohesion.9 = lsm_l_cohesion(vosssurface9, directions=8)
PatchCohesion.9 <- PatchCohesion.9$value


# Data Frame to Plot
LandscapeCohesion <- data.frame(
  HValue = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9),
  CohesionValues = c(PatchCohesion.1, PatchCohesion.2, PatchCohesion.3,
                 PatchCohesion.4, PatchCohesion.5, PatchCohesion.6,
                 PatchCohesion.7, PatchCohesion.8, PatchCohesion.9)
)

# Create Line of Best Fit For Inclusion on Figure
CohesionModel <- lm(CohesionValues ~ HValue, data = LandscapeCohesion)
CohesionSlope <- coef(CohesionModel)["HValue"]
CohesionIntercept <- coef(CohesionModel)["(Intercept)"]
CohesionLineLabel <- paste("y =", round(CohesionIntercept, 2), "+", round(CohesionSlope, 2), "H")

# Adjust the y position for the annotation to be within the visible range
adjusted_y_position_for_label <- 99.5 

# Plot
ggplot(LandscapeCohesion, aes(x = HValue, y = CohesionValues)) +
  geom_point() +  
  geom_smooth(method = "lm", se = TRUE, color = "red") + 
  annotate("text", x = 0.5, y = adjusted_y_position_for_label, label = CohesionLineLabel, hjust = 0, vjust = 1, color = "red") +
  coord_cartesian(ylim = c(99, max(LandscapeCohesion$CohesionValues, na.rm = TRUE))) + 
  theme_minimal() +
  labs(title = "Landscape Patch Cohesion vs H Value",
       x = "H",
       y = "Patch Cohesion")

################################################################################

#Landscape Core Area Means
CoreAreaMean.1 = lsm_l_cai_mn(vosssurface1, directions=8)
CoreAreaMean.1 <- CoreAreaMean.1$value
CoreAreaMean.2 = lsm_l_cai_mn(vosssurface2, directions=8)
CoreAreaMean.2 <- CoreAreaMean.2$value
CoreAreaMean.3 = lsm_l_cai_mn(vosssurface3, directions=8)
CoreAreaMean.3 <- CoreAreaMean.3$value
CoreAreaMean.4 = lsm_l_cai_mn(vosssurface4, directions=8)
CoreAreaMean.4 <- CoreAreaMean.4$value
CoreAreaMean.5 = lsm_l_cai_mn(vosssurface5, directions=8)
CoreAreaMean.5 <- CoreAreaMean.5$value
CoreAreaMean.6 = lsm_l_cai_mn(vosssurface6, directions=8)
CoreAreaMean.6 <- CoreAreaMean.6$value
CoreAreaMean.7 = lsm_l_cai_mn(vosssurface7, directions=8)
CoreAreaMean.7 <- CoreAreaMean.7$value
CoreAreaMean.8 = lsm_l_cai_mn(vosssurface8, directions=8)
CoreAreaMean.8 <- CoreAreaMean.8$value
CoreAreaMean.9 = lsm_l_cai_mn(vosssurface9, directions=8)
CoreAreaMean.9 <- CoreAreaMean.9$value

# Data Frame to Plot
LandscapeCoreAreaMean <- data.frame(
  HValue = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9),
  CoreAreaValues = c(CoreAreaMean.1, CoreAreaMean.2, CoreAreaMean.3,
                 CoreAreaMean.4, CoreAreaMean.5, CoreAreaMean.6,
                 CoreAreaMean.7, CoreAreaMean.8, CoreAreaMean.9)
)

# Create Line of Best Fit For Inclusion on Figure
CoreAreaMeanModel <- lm(CoreAreaValues ~ HValue, data = LandscapeCoreAreaMean)
CoreAreaMeanSlope <- coef(CoreAreaMeanModel)["HValue"]
CoreAreaMeanIntercept <- coef(CoreAreaMeanModel)["(Intercept)"]
CoreAreaMeanLineLabel <- paste("y =", round(CoreAreaMeanIntercept, 2), "+", round(CoreAreaMeanSlope, 2), "H")

# Plot
ggplot(LandscapeCoreAreaMean, aes(x = HValue, y = CoreAreaValues)) +
  geom_point() +  
  geom_smooth(method = "lm", se = TRUE, color = "red") + 
  annotate("text", x = 0.5, y = max(LandscapeCoreAreaMean$CoreAreaValues, na.rm = TRUE), label = CoreAreaMeanLineLabel, hjust = 0, vjust = 1, color = "red") +
  theme_minimal() +
  labs(title = "Landscape Core Area Mean vs H Value",
       x = "H",
       y = "Core Area Mean")
################################################################################

# Landscape Number of Core Areas
NumCoreAreas.1 = lsm_l_ndca(vosssurface1, directions=8)
NumCoreAreas.1 <- NumCoreAreas.1$value
NumCoreAreas.2 = lsm_l_ndca(vosssurface2, directions=8)
NumCoreAreas.2 <- NumCoreAreas.2$value
NumCoreAreas.3 = lsm_l_ndca(vosssurface3, directions=8)
NumCoreAreas.3 <- NumCoreAreas.3$value
NumCoreAreas.4 = lsm_l_ndca(vosssurface4, directions=8)
NumCoreAreas.4 <- NumCoreAreas.4$value
NumCoreAreas.5 = lsm_l_ndca(vosssurface5, directions=8)
NumCoreAreas.5 <- NumCoreAreas.5$value
NumCoreAreas.6 = lsm_l_ndca(vosssurface6, directions=8)
NumCoreAreas.6 <- NumCoreAreas.6$value
NumCoreAreas.7 = lsm_l_ndca(vosssurface7, directions=8)
NumCoreAreas.7 <- NumCoreAreas.7$value
NumCoreAreas.8 = lsm_l_ndca(vosssurface8, directions=8)
NumCoreAreas.8 <- NumCoreAreas.8$value
NumCoreAreas.9 = lsm_l_ndca(vosssurface9, directions=8)
NumCoreAreas.9 <- NumCoreAreas.9$value

# Data Frame to Plot
LandscapeNumCoreAreas <- data.frame(
  HValue = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9),
  CoreAreas = c(NumCoreAreas.1, NumCoreAreas.2, NumCoreAreas.3,
                 NumCoreAreas.4, NumCoreAreas.5, NumCoreAreas.6,
                 NumCoreAreas.7, NumCoreAreas.8, NumCoreAreas.9)
)

# Create Line of Best Fit For Inclusion on Figure
NumCoreAreasmodel <- lm(CoreAreas ~ HValue, data = LandscapeNumCoreAreas)
NumCoreAreasmodelslope <- coef(NumCoreAreasmodel)["HValue"]
NumCoreAreasmodelintercept <- coef(NumCoreAreasmodel)["(Intercept)"]
NumCoreAreasmodelline_label <- paste("y =", round(NumCoreAreasmodelintercept, 2), round(NumCoreAreasmodelslope, 2), "H")

# Plot
ggplot(LandscapeNumCoreAreas, aes(x = HValue, y = CoreAreas)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  annotate("text", x = 0.5, y = max(LandscapeNumCoreAreas$CoreAreas), label = NumCoreAreasmodelline_label, hjust = 0, vjust = 1, color = "red") +
  theme_minimal() +
  labs(title = "Landscape Core Areas vs H Values",
       x = "H",
       y = "Number of Disjunct Core Areas")


```

**ANSWER:**
*Landscape Mean of Core Area Index:* 
This metric was chosen because increased H values indicated forest cover becoming more clumped. Consequently, it is anticipated a rise in core area and a decline in total edge with increasing clumpiness, supported by the slope in the plot, where H values are on the x-axis and the Landscape Mean of Core Area Index is on the y-axis.

*Landscape Number of Core Areas:* Concerning the Core Area Index, this metric was chosen because it is expected a decrease in total core areas as forest cover became more aggregated, indicating an inverse relationship with the Mean of Core Area Index. This prediction was supported by the observed slope in the plot, showing a decrease in the number of disjunctive core areas as H values increased across the landscape.

*Landscape Patch Cohesion Index:* This metric was chosen because as H increases, forest cover appeared more clumped. Consequently, the expected Patch Cohesion Index, a measure of aggregation, to increase with higher H values. Plotting H against the Landscape Patch Cohesion Index supported the prediction with a positive slope. However, the Patch Cohesion Index showed minimal variation, ranging from approximately 99.05 to 99.5.
